---
title: "Bayesian Data Analysis of an Uber Driver's Optimal Daily Circuit"
author: "Krishnan Raman"
date: "12/3/2020"
output: 
  beamer_presentation: 
    theme: Darmstadt
    colortheme: seagull
    fonttheme: default
    fig_caption: yes
---

Contents
========================================================

- Uber Graphs & Subgraphs
- What's an Optimal Daily Circuit ?
- Finding the Optimal Daily Circuit
- Optimal Circuit Summary Statistics
- Posterior Mean of the fare from the Optimal Daily Circuit ( Gaussian Conjugate Prior )
- Posterior Mean of the fare from the Optimal Daily Circuit ( Hierarchical Models, Stan/MCMC )
- Posterior Mean of the fare from a Random Daily Circuit ( Hierarchical Models, Stan/MCMC )
- Comparing Two Optimal circuits ( Confidence Distribution, p-value function )


Uber Dataset
========================================================
- 29 million rows
```
  orig  dest  pickup_datetime   distance duration  fare
1  7C    6A 2014-09-01 09:00:00     4.25   15.183 15.30
2  7B    15 2014-09-01 18:00:00    10.17   34.083 32.28
3  11    2A 2014-09-01 17:00:00     4.02   17.100 15.57
4  3B    4A 2014-09-01 13:00:00     1.46    6.533  8.00
5  2A    10 2014-09-01 14:00:00     8.31   26.283 26.29
6  5B    4C 2014-09-01 12:00:00     1.04    8.583  8.00
```

Uber Traffic as a Complete Graph
===============================================
```{r, echo=FALSE,out.height="80%"}
suppressMessages(library(igraph))
# make graph of original dataset
all_edges <- readRDS("all_edges.Rda")
n<- dim(all_edges)[1]
e1<-matrix(NA,nrow=n,ncol=3)
for(i in 1:n) {
  e1[i,] = c(all_edges[[i,1]],all_edges[[i,2]],all_edges[[i,3]]) # from, to, weight
}
g<-graph_from_edgelist(e1[,1:2], directed=FALSE)
plot(g, layout=layout.circle, main=paste("Uber as a graph, with ", length(V(g)), "vertices,", length(E(g)), "edges."))
```

Subgraph induced by Top-100 most lucrative edges.
===============================================
```{r, echo=FALSE,out.height="80%"}
# make graph of original dataset
suppressMessages(library(igraph))
setwd("~/R/Stat 695/Project/uber/")
topk_edges <- readRDS("topk.Rda")
n<- dim(topk_edges)[1]
edgematrix<-matrix(NA,nrow=n,ncol=3)
for(i in 1:n) {
  edgematrix[i,] = c(topk_edges[[i,1]],topk_edges[[i,2]],topk_edges[[i,3]])
}
g<-graph_from_edgelist(edgematrix[,1:2], directed=FALSE)
plot(g, layout=layout.circle, main=paste("Uber top-100 subgraph, Vertices:", length(V(g)), "Edges:", length(E(g))))
```

What's an Optimal Daily Circuit ?
============================================
- Circuit: Collection of edges
- Want a closed circuit aka Cycle or loop
- Want a "simple path" : each node visited only once

## Optimal Circuit
- Circuit that makes the most money for the Uber Driver

## Optimal Daily Circuit
- Driver works 9AM-5PM: 8 HOURS
- Make 1 Uber trip per hour
- Want an 8-cycle aka Simple Path with 8 Nodes aka 8-gon

Does a $60 per edge sparse graph contain an 8-gon ?
============================================
```{r, echo=FALSE,out.height="80%"}
highfare_matrix = edgematrix[as.numeric(edgematrix[,3]) > 60,]
highfare_weights = round(as.numeric(highfare_matrix[,3]))
g2<-graph_from_edgelist(highfare_matrix[,1:2], directed=FALSE)
plot(g2,edge.label= highfare_weights, layout=layout.circle)
```

The Optimal 8-gon 
==================
```{r, echo=FALSE,out.height="80%"}
for(fares in seq(60,45,-5)) {
  highfare_matrix = edgematrix[as.numeric(edgematrix[,3]) > fares,]
  g2<-graph_from_edgelist(highfare_matrix[,1:2], directed=FALSE)
  lengths = c()

  for (v in V(g2)$name) {
    res = all_simple_paths(g2,v)
   lengths = c(lengths, max(sapply(1:length(res), function(j) { length(res[[j]]) })))
  }
  cat(sprintf("Fare: $%.0f, Longest path: %.0f, Vertices:%.0f Edges %.0f\n", fares, max(lengths), length(V(g2)),length(E(g2))))
}

# visualize the graph with simple path of length 8
highfare_matrix = edgematrix[as.numeric(edgematrix[,3]) > 46,]
highfare_weights = round(as.numeric(highfare_matrix[,3]))
g3<-graph_from_edgelist(highfare_matrix[,1:2], directed=FALSE)
paths_of_length_8 = c()
found=FALSE
for (src in V(g3)$name) {
    res = all_simple_paths(g3,src)
    for(i in 1:length(res)) {
      l = length(res[[i]])
      last = res[[i]][l]$name
      d = distances(g3,v=last,to=src)
      if (l == 8 & d[1] == 1) {
          paths_of_length_8 = c(paths_of_length_8, res[[i]])
          #print(res[[i]])
          found = TRUE
          break
      }
    }
    if(found) break
}
plot(g3, mark.groups=names(paths_of_length_8[1:8]), mark.col="red")
```

How much does the 8-gon Uber driver make ?
============================================
```{r, echo=FALSE,out.height="80%"}
getidx = function(src,dest) {
  #cat(paste(src,",",dest, "\n"))
  srcidx = which(topk_edges == src, arr.ind=TRUE)
  destidx = which(topk_edges == dest, arr.ind=TRUE)
 
  r = intersect(srcidx[,1], destidx[,1])[1]
  #print(srcidx)
  #print(destidx)
  #print(r)
  return(r)
}

get_fare_dist = function(src,dest) {
   r = getidx(src,dest)
   return(topk_edges[[r,4]])
}

get_median_fare=function(src,dest) {
  r = getidx(src,dest)
  if (topk_edges[[r,1]] == src & topk_edges[[r,2]] == dest) {
     return(topk_edges[[r,3]])  
  } else if (topk_edges[[r,2]] == src & topk_edges[[r,1]] == dest) {
    return(topk_edges[[r,3]])
  } else return(-1)
}


vlist = c()
for(i in 1:7) {
  vlist = c(vlist, names(paths_of_length_8[i]), names(paths_of_length_8[i+1]))
}
vlist = c(vlist, names(paths_of_length_8[8]), names(paths_of_length_8[1]))
edge_wt = numeric(8)
for(i in 0:7) {
  f = 2*i + 1
  t = f+1
  sf = vlist[f]
  dt = vlist[t]
  edge_wt[i+1] = get_median_fare(sf,dt)
}

plot(make_undirected_graph(vlist),edge.color="red", layout=layout.circle, edge.label=round(edge_wt))
```

Optimal Circuit Summary Statistics
============================================
```{r, echo=FALSE,out.height="80%"}
stats_df = data.frame()
ckt_dist = numeric(8)
for(i in 0:7) {
  f = 2*i + 1
  t = f+1
  sf = vlist[f]
  dt = vlist[t]
  dist = get_fare_dist(sf,dt)
  ckt_dist[(i+1)] = getidx(sf,dt)
  stats_df[(i+1),1:2] = c(sf,dt)
  stats_df[(i+1),3:6] = round(c(mean(dist), median(dist), sd(dist), var(dist)),2)
}
colnames(stats_df) <- c("Src", "Dest", "Mean", "Median", "Sigma", "Var")
print(stats_df)
```

Optimal Circuit: Edge Distributions
============================================
```{r, echo=FALSE,out.height="80%"}
#obtain n samples from the circuit just once
n = 1000
get_circuit_sample <- function() {
  return(sum(sapply(1:8, function(i) { 
    r = ckt_dist[i]
    dist = topk_edges[[r,4]]  
    sample(dist,1) 
  })))
}

y_i = sapply(1:n, function(i){get_circuit_sample()})

par(mfrow=c(2,4))
for(i in 1:8) {
  r = ckt_dist[i]
  dist = topk_edges[[r,4]]  
  hist(dist,breaks=20, main=paste("Edge ", vlist[2*(i-1)+1], "->", vlist[2*i]))
}
```

Posterior Mean of the fare from the Optimal Daily Circuit
==========================================================
- Edge Fare $X_i$ = Observed fares for edge $X_i$ over a year.        
- Circuit Fare $Y = \sum X_i$, i=1:8, Y is unobserved.   

## Model 1 : Gaussian Conjugate Prior
- $X_i$ approx Gaussian, sample mean \$50, sample sd \$7   
- We attribute a Gaussian prior $Y \sim N(50*8, 8*(7^2))$
- Posterior Dist of Y is product of Prior & Likelihood   
- Gaussian Conjugate prior yields Gaussian posterior

## Model 2 : Hierarchical Model   
- $X_i \sim N(\mu_i,\sigma)$
- $\mu_i \sim N(\mu,\phi)$
- $Y \sim N(8\mu,\sqrt{8}\sigma)$
- Estimate posterior mean of Y via MCMC

Posterior Mean of the fare from the Optimal Daily Circuit
==========================================================
```{r, echo=FALSE,out.height="80%"}
n = 50
edge_fares = sapply(1:8, function(i) { 
    r = ckt_dist[i]
    dist = topk_edges[[r,4]]  
    mean(dist[1:n]) 
  })

# data
sigma_y = n*sum(edge_fares)
sigmasq = 392 # 6sigma = 42, sigma^2 = 49, 8*sigma^2 = 392

# prior
mu0 = 400 #$50 per edge times 8
tau0sq = 392 

# posterior
denom = (1/tau0sq) + (n/sigmasq)
mu = (mu0/tau0sq + sigma_y/sigmasq)/denom
sigmasq_post = 1/denom

#plotting vars
x = seq(mu0-3*sqrt(tau0sq),mu0+3*sqrt(tau0sq),.1)
y = dnorm(x,mean=mu,sd=sqrt(sigmasq_post))
yp = dnorm(x,mean=mu0,sd=sqrt(tau0sq))

# Gaussian prior
plot(x,y, main=paste("Posterior Mean:",round(mu,2)), col="blue", "l")
lines(x,yp,col="red")
abline(v=mu0, col="red")
```

Fare Distribution: Optimal Daily Circuit
========================================
```{r, echo=FALSE,out.height="80%"}

scode <- "
data {
  int<lower=0> N;
  vector[N] y;
  vector[N] x1;
  vector[N] x2;
  vector[N] x3;
  vector[N] x4;
  vector[N] x5;
  vector[N] x6;
  vector[N] x7;
  vector[N] x8;
}

parameters {
  real mu1;
  real mu2;
  real mu3;
  real mu4;
  real mu5;
  real mu6;
  real mu7;
  real mu8;
  real mu;
  real<lower=0> phi; 
  real<lower=0> sigma;
}

model {
  y ~ normal(mu*8, sqrt(8)*sigma);
  x1 ~ normal(mu1, sigma);
  x2 ~ normal(mu2, sigma);
  x3 ~ normal(mu3, sigma);
  x4 ~ normal(mu4, sigma);
  x5 ~ normal(mu5, sigma);
  x6 ~ normal(mu6, sigma);
  x7 ~ normal(mu7, sigma);
  x8 ~ normal(mu8, sigma);
  mu1 ~ normal(mu, phi);
  mu2 ~normal(mu,phi);
  mu3 ~normal(mu,phi);
  mu4 ~ normal(mu, phi);
  mu5 ~normal(mu,phi);
  mu6 ~normal(mu,phi);
  mu7 ~ normal(mu, phi);
  mu8 ~normal(mu,phi);
}
"
find_edge_dist = function(i,N) {
  return(sample(topk_edges[[ckt_dist[i],4]],N,FALSE))
}

N=50
x = matrix(0,8,N)
y = numeric(50)
for( i in 1:8) {
  x[i,] = find_edge_dist(i,N)
  y = y + x[i,]
}
hist(y, main="Fare distribution of Optimal Daily Circuit")
```

Posterior Mean of the fare from the Optimal Daily Circuit (Hierarchical Models, Stan/MCMC)
===========================================================
```{r, echo=FALSE,out.height="80%",error=FALSE, warning=FALSE,message=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

data = list(N=N,x1=x[1,],x2=x[2,],x3=x[3,],x4=x[4,],x5=x[5,],x6=x[6,],x7=x[7,],x8=x[8,],y=y)

sink("/dev/null")  
suppressMessages(library(rstan))
rstan_options(auto_write = FALSE)
options(mc.cores = parallel::detectCores())
fit <- stan(model_code = scode, data = data, iter=500, warmup=50,chains=1)
sink()
```
```{r, echo=FALSE,out.height="80%",error=FALSE, warning=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
s<-summary(fit, probs = c(0.5))
optimal_fare<-8*s$summary[9]
suppressMessages(plot(fit, pars=c("mu","phi","mu1","mu2","mu3","mu4","mu5","mu6","mu7","mu8")))
```


Posterior Mean of the fare from the Optimal Daily Circuit (Hierarchical Models, Stan/MCMC)
===========================================================
```{r, echo=FALSE,out.height="80%"}
cat(sprintf("Mean Fare from Optimal Daily Circuit $%.2f\n", optimal_fare))
traceplot(fit,pars="mu")
```

Posterior Mean of the fare from a Random Daily Circuit (Hierarchical Models, Stan/MCMC)
===========================================================
```{r, echo=FALSE,out.height="80%", error=FALSE, warning=FALSE, message=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

N=50
x = matrix(0,8,N)
ry = numeric(50)
set.seed(12343)
random_paths = sample(1:812,8)
find_path_dist = function(i,N) {
  return(sample(all_edges[[i,4]],N,FALSE))
}
for( i in 1:8) {
  x[i,] = find_path_dist(random_paths[i],N)
  ry = ry + x[i,]
}

data = list(N=N,x1=x[1,],x2=x[2,],x3=x[3,],x4=x[4,],x5=x[5,],x6=x[6,],x7=x[7,],x8=x[8,],y=ry)

sink("/dev/null")  
suppressMessages(library(rstan))
rstan_options(auto_write = FALSE)
options(mc.cores = parallel::detectCores())
rfit <- stan(model_code = scode, data = data, iter=500, warmup=50,chains=1)
sink()
```

```{r, echo=FALSE,out.height="80%", error=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rs<-summary(rfit, probs = c(0.5))
random_fare<-8*rs$summary[9]
daily_loss<- optimal_fare - random_fare

suppressMessages(plot(rfit, pars=c("mu","phi","mu1","mu2","mu3","mu4","mu5","mu6","mu7","mu8")))
```

Posterior Mean of the fare from a Random Daily Circuit (Hierarchical Models, Stan/MCMC)
===========================================================
```{r, echo=FALSE,out.height="80%"}
cat(sprintf("Mean Fare from Random Daily Circuit $%.2f\n", random_fare))
cat(sprintf("|Optimal-Random| Median Loss: $%.2f\n", daily_loss))
traceplot(fit,pars="mu")
```

Comparing two Uber circuits via Confidence Distributions
============================================
- Professor Don Rubin: "Fisher Randomization Test is a Stochastic Proof-By-Contradiction"
- FRT = distribution-free test to compare two (multimodal) Uber circuits
- Fisher Sharp Null compares individual potential outcomes $Y_i(1) \ vs \ Y_i(0)$ for every observation.
- An assignment is a Boolean vector over two weeks. 
- An assignment simply means on the given day, the Uber driver drove Circuit c2.
- A non-assignment means the wages would come from Ckt c1 aka Null Hypothesis. 
- Compute the biweekly average and compare with the null hypothesis biweekly average
- This comparison is a simple difference test statistic.


Comparing two Uber circuits via Confidence Distributions
============================================
```{r, echo=FALSE,out.height="80%"}
n<-14 # 2 weeks
options(digits=4)
c1<- sample(y,n)
c2<- sample(ry,n)

outersect <- function(x, y) {
  sort(c(x[!x%in%y],
         y[!y%in%x]))
}

 ".combinadic" <- function(n, r, i) {
     
     if(i < 1 | i > choose(n,r)) stop("'i' must be 0 < i <= n!/(n-r)!")
     
     largestV <- function(n, r, i) {
         v <- n                                  # Adjusted for one-based indexing
         while(choose(v,r) >= i) v <- v-1        # Adjusted for one-based indexing
         return(v)
     }
     
     res <- rep(NA,r)
     for(j in 1:r) {
         res[j] <- largestV(n,r,i)
         i <- i-choose(res[j],r)
         n <- res[j]
         r <- r-1
     }
     res <- res + 1
     return(res)
}
n=14
r<-(n/2)
len <- choose(n,r)
t <- numeric(len)
for(i in 1:len) {
  assignment<-.combinadic(n,r,i)
  y_obs <- c1
  y_obs[assignment] <- c2[assignment]
  not_assigned <- outersect(1:n, assignment)
  # want Ybar(1) - Ybar(0) on the observed
  diff <- mean(y_obs[assignment]) - mean(y_obs[not_assigned])
  t[i] <- diff
}

m<- round(max(abs(min(t)), abs(max(t))))
theta <- seq(-m,m,1)
nt <- length(t)
pvalues<-c()
for(mytheta in theta) {
  pval<- length(t[t < mytheta])/nt
  pvalues<-c(pvalues, pval)
}

par(mfrow=c(1,2))
hist(t, main="Difference b/w wages of 2 ckts")
plot(theta,pvalues, main="p-value function", 'l')
abline(h=0.05,col='red')
abline(h=0.95,col='red')
```